Review Paper Draft

### 2020-06-18 lmm-genetic-review.tex
Edits to Introduction

\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{float}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{url}
\allowdisplaybreaks
%\usepackage{biblatex}
%\AtEveryCitekey{%
%  \clearfield{note}%
%}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\usetikzlibrary{arrows}
\providecommand{\note}[1]{\textcolor{red}{#1}}
\providecommand{\anna}[1]{\textcolor{blue}{#1}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bbetaT}{\widetilde{\boldsymbol{\beta}}}
\newcommand{\bbetaHat}{\widehat{\boldsymbol{\beta}}}
\newcommand{\bgamma}{\boldsymbol{\gamma}}
\newcommand{\beps}{\boldsymbol{\varepsilon}}
\newcommand{\bepsT}{\widetilde{\boldsymbol{\varepsilon}}}
\newcommand{\bX}{\textbf{X}}
\newcommand{\bXT}{\widetilde{\textbf{X}}}
\newcommand{\bZ}{\textbf{Z}}
\newcommand{\bZT}{\widetilde{\textbf{Z}}}
\newcommand{\by}{\textbf{y}}
\newcommand{\byT}{\widetilde{\textbf{y}}}
\newcommand{\bD}{\textbf{D}}
\newcommand{\bK}{\textbf{K}}
\newcommand{\bI}{\textbf{I}}
\newcommand{\bV}{\textbf{V}}
\newcommand{\bU}{\textbf{U}}
\newcommand{\bR}{\textbf{R}}
\newcommand{\bW}{\textbf{W}} 
\newcommand{\sigmaee}{\sigma^2_e}
\newcommand{\sigmagg}{\sigma^2_g}

@@

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to confounding due to unobserved factors. This poses a challenge to the detection of multivariate associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. Despite extensive discussion and simulation studies of penalized LMMs in the literature, the non-genetic mechanism by which population stratification confounds the phenotype-genotype relationship is rarely discussed or treated explicitly. As such, the ability of LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Considering this, we evaluate the performance of several penalized regression methods in terms of MSE in presence of a number of confounding structures.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

>>>>

\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{float}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{url}
\allowdisplaybreaks
%\usepackage{biblatex}
%\AtEveryCitekey{%
%  \clearfield{note}%
%}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\usetikzlibrary{arrows}
\providecommand{\pb}[1]{\textcolor{red}{#1}}
\providecommand{\anna}[1]{\textcolor{blue}{#1}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bbetaT}{\widetilde{\boldsymbol{\beta}}}
\newcommand{\bbetaHat}{\widehat{\boldsymbol{\beta}}}
\newcommand{\bgamma}{\boldsymbol{\gamma}}
\newcommand{\beps}{\boldsymbol{\varepsilon}}
\newcommand{\bepsT}{\widetilde{\boldsymbol{\varepsilon}}}
\newcommand{\bX}{\textbf{X}}
\newcommand{\bXT}{\widetilde{\textbf{X}}}
\newcommand{\bZ}{\textbf{Z}}
\newcommand{\bZT}{\widetilde{\textbf{Z}}}
\newcommand{\by}{\textbf{y}}
\newcommand{\byT}{\widetilde{\textbf{y}}}
\newcommand{\bD}{\textbf{D}}
\newcommand{\bK}{\textbf{K}}
\newcommand{\bI}{\textbf{I}}
\newcommand{\bV}{\textbf{V}}
\newcommand{\bU}{\textbf{U}}
\newcommand{\bR}{\textbf{R}}
\newcommand{\bW}{\textbf{W}} 
\newcommand{\sigmaee}{\sigma^2_e}
\newcommand{\sigmagg}{\sigma^2_g}

@@

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to confounding due to unobserved factors \pb{``Unobserved factors'' is a bit vague -- think we need to work ``environment'' into this first sentence}. This poses a challenge to the detection of multivariate associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. Despite extensive discussion and simulation studies of penalized LMMs in the literature \pb{is the literature on penalized LMMs really ``extensive''?  I would say that the literature on penalized regression is extensive, and on LMMs is extensive, but not actually that much on penalized LMMs -- hence the need for this review}, the non-genetic mechanism by which population stratification confounds the phenotype-genotype relationship is rarely discussed or treated explicitly. As such, the ability of LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Considering this, we evaluate the performance of several penalized regression methods in terms of MSE in presence of a number of confounding structures.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

### 2020-06-18 main.tex
Experimental push
Edits to Introduction

Genetic variants are often correlated with each other, meaning that univariate testing of genetic variants may result in false positives and biased effect estimates. This occurs when the variant being tested is correlated with other variants, so that the effects of these correlated variants are erroneously attributed to the variant in question. One method for analyzing data in the presence of LD is pruning, where only uncorrelated SNPs are used in modeling or testing procedures. Pruning limits analysis to SNPs which are representative of a particular region, and may result in improved predictive accuracy. However, pruning is limited in that it can only identify SNPs in LD with the true causal SNP. In addition, it has been shown that a worst-case scenario, pruning algorithms may not select any representative SNPs in a particular region of the genome \cite{prive2018efficient}.

An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, such methods rely on data where the number of observations is less than the number of of features. This is often infeasible in genetic studies where the number of features frequently exceeds the number of observations. Penalized regression methods make such problems tractable, by constraining the solution space of coefficient estimates.

The Lasso \cite{tibshirani1996regression} is a widely used penalized regression method. It relies on an assumption of sparsity, which allows for computational speedup while performing coefficient estimation and variable selection simultaneously. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

Lasso penalized regression models assume independent observations and uncorrelated errors. In the presence of population structure, some samples are more correlated than others, violating the independent errors assumption. If left unaccounted for, this presence of such dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \cite{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, there remain open and relevant problems in analyzing structured genetic data \cite{lawson2019population, barton2019population}.

@@

Using the spectral decomposition of $\bK$ described in equation \ref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorizations

>>>>

Genetic variants are often correlated with each other, meaning that univariate testing of genetic variants may result in false positives and biased effect estimates. This occurs when the variant being tested is correlated with other variants, so that the effects of these correlated variants are erroneously attributed to the variant in question. One method for analyzing data in the presence of LD \pb{LD needs to be defined} is pruning, where only uncorrelated SNPs are used in modeling or testing procedures. Pruning limits analysis to SNPs which are representative of a particular region, and may result in improved predictive accuracy \pb{is predictive accuracy the main reason for pruning?  Why does it improve prediction?}. However, pruning is limited in that it can only identify SNPs in LD with the true causal SNP \pb{I don't understand this sentence; wouldn't this limitation apply to all methods?}. In addition, it has been shown that in the worst-case scenario, pruning algorithms may not select any representative SNPs in a particular region of the genome \cite{prive2018efficient}.

An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, such methods rely on data where the number of observations is less than the number of features. This is often infeasible in genetic studies where the number of features typically exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates.

The Lasso \cite{tibshirani1996regression} is a widely used penalized regression method. It relies on an assumption of sparsity, which means that coefficient estimation and variable selection can be accomplished simultaneously, and also provides considerable computational advantages, meaning that it scales up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

Lasso penalized regression models typically assume independent observations and uncorrelated errors. In the presence of population structure, some samples are more correlated than others, violating the independent errors assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \cite{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, there remain open and relevant problems \pb{this is vague} in analyzing structured genetic data \cite{lawson2019population, barton2019population}.

@@

Using the spectral decomposition of $\bK$ described in equation \ref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization \pb{ONLY WORKS IF $\bX$ IS FULL RANK}
